---
title: "covid-19"
author: "Alexey Sokolov"
date: '2022-07-18'
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r import_library, include=FALSE}
library(tidyverse)
library(lubridate)
```

### Data source and overview
Data was downloaded from this website - https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series
Data represents daily time series of COVID-19 cases including confirmed,
recovered and deaths

## Import data
here we just use read_csv data function that takes URL as a parameter

```{r import_data}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_global.csv",
               "time_series_covid19_deaths_global.csv",
               "time_series_covid19_confirmed_US.csv",
               "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in, file_names)

global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
us_cases <- read_csv(urls[3])
us_deaths <- read_csv(urls[4])
```


## Tidy data
here we "lengthen" data (decrease the number of columns and increase the number
of rows). Previously we had column for every possible date and now we put this
information in rows using pivot_longer function. Also we remove Lat and Long as
we won't use it in further analysis. And finally we convert date to date format
using mdy function

```{r tidy_data}
global_cases <- global_cases %>% 
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), 
               names_to = "date", 
               values_to = "cases") %>% 
  select(-c(Lat, Long))

global_deaths <- global_deaths %>% 
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long), 
               names_to = "date", 
               values_to = "deaths") %>% 
  select(-c(Lat, Long))


us_cases <- us_cases %>%
  pivot_longer(cols = - (UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

us_deaths <- us_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
```

## Transform data
Here we joint global_cases and global_deaths datasets using full_join column and
filter all rows that don't have any cases using filter function. We also add
information about population size using uid variable.
For US data we just join us_cases with us_deaths and store new dataset in us 
variable
```{r transform_data}
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>%
  mutate(date = mdy(date))
global <- global %>% filter(cases > 0)
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, Population,
         Combined_Key)

us <- us_cases %>%
  full_join(us_deaths)
```

## Summary of data
here we show summary of data using summary function
```{r summary_data}
summary(global)
summary(us)
```


## Visualizing data

for this, let's summarize number of cases, deaths, population and introduce new
column - deaths per million. After that let's plot number of cases and deaths in
USA and New York. From the plots bellow it's clear that these two characteristics
are correlated.
```{r visualize_data}
us_by_state <- us %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
           Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill,
         Population) %>%
  ungroup()

us_totals <- us_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

us_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL)

state <- "New York"
us_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y = NULL)

max(us_totals$date)
max(us_totals$deaths)
```

## Analyzing data
Next let's introduce two new variables: new_cases and new_deaths and plot them in
USA and New York.
```{r analyze_data}
us_by_state <- us_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

us_totals <- us_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
tail(us_totals %>% select(new_cases, new_deaths, everything()))

us_totals %>%
  ggplot(aes(x = date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) + 
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL)

state <- "New York"
us_by_state %>%
  filter(Province_State == state) %>%
  ggplot(aes(x = date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) + 
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y = NULL)

us_state_totals <- us_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000 * cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)

us_state_totals %>%
  slice_min(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

us_state_totals %>%
  slice_max(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
```

## Model data in the USA
Here we will use linear model to build a model of a dependency between deaths per
thousand and cases per thousand. When we have model, we can then plot it against
real data that we have.
```{r model_data}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = us_state_totals)
summary(mod)
us_state_totals %>% slice_min(cases_per_thou)
us_state_totals %>% slice_max(cases_per_thou)
us_state_totals %>% mutate(pred = predict(mod))

us_tot_w_pred <- us_state_totals %>% mutate(pred = predict(mod))
us_tot_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```

### Analyze worldwide data
let's plot a graph of global deaths and global cases and then a graph of cases
and deaths in China. For China I can clearly see huge increase in number of cases
for 2022.
```{r global_analysis}
global %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in the World", y = NULL)

country <- "China"
global %>%
  filter(Country_Region == country) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", country), y = NULL)
```

### Linear model for global data
Next let's build linear model for global data and try to find a model for a number
of cases and number of deaths dependency.
```{r global_lm}
mod <- lm(cases ~ deaths, data = global)
summary(mod)
```

### Bias analysis
As any human I have lots of biases. For example I can have a biases towards 
some specific countries. I think there are two possible solutions to mitigate that:

1. Analyze data in all possible combinations with all possible types of 
visualizations. That's ideal solution but quite often it's just not feasible
2. Work only with unpersonalized data, for example instead of Categorical 
country_region columns it's possible to use some numbers instead 
(for example 0 - Afghanistan and 1 - China, etc...)